---
title: "HMM search output analyses with avg e-val:s"
author: "LJM"
date: "`r Sys.Date()`"
link-citations: true
bibliography: bibliography.bib
biblio-style: "alpha" #https://www.overleaf.com/learn/latex/Bibtex_bibliography_styles
#csl: "../uppsala-universitet-institutionen-for-biologisk-grundutbildning.csl"
documentclass: report
output:
  bookdown::html_document2:
    highlight: tango
    theme: journal
    split_by: none # only generate a single output page
    self_contained: TRUE
    toc: yes
    toc_float: 
      collapsed: TRUE
      smooth_scroll: TRUE
      print: FALSE
    code_folding: show
    number_sections: TRUE
    code_download: TRUE
  html_document: 
    toc: TRUE
    toc_float: TRUE
    code_folding: show
    highlight: tango
    number_sections: TRUE
    self_contained: TRUE
    smart: TRUE # To be figured out when this is useful
    theme: journal
  pdf_document:
    toc: TRUE
    toc_depth: 2
    highlight: tango # This appears to be the same as default
    number_sections: TRUE
  bookdown::html_document2:
    highlight: tango
    theme: journal
    split_by: none # only generate a single output page
    self_contained: TRUE
    toc: yes
    toc_float: 
      collapsed: TRUE
      smooth_scroll: TRUE
      print: FALSE
    code_folding: show
    number_sections: TRUE
    code_download: TRUE
---

# Load libraries

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(vegan)
library(ggbiplot)
```

# Read intermediary `csv` file

In the intermediary `csv` file the $-log_{10}(\text{e-val})$:s are the average e-values of all matches with same HMM profiles to each fOTU.

```{r eval=FALSE, include=TRUE}
columns <- paste("c", strrep("d",8315), sep = "")
fOTUsHMM <- read_csv(file = "../analyses/HMMsearch/avg_fOTUsHMM.csv", col_types = columns)
```

# Exchange NA, Inf and -Inf with 0

```{r}
is.na(fOTUsHMM) <- sapply(fOTUsHMM, is.infinite)
fOTUsHMM[is.na(fOTUsHMM)] <- 0
```

# Move fOTU names to row names and execute PCA 

```{r}
fOTUsHMM_pca <- fOTUsHMM %>% 
  column_to_rownames(var = "fOTU_name") %>%
  prcomp(., 
         center = TRUE)
```

# Visualise the results

The following graph depicts how large percentage of bins in fOTU have at least one hit from viral HMM profiles with e-value less than 0.01.

```{r}
file_name <- "../visualisations/HMMsearch/ggbiplot_avg_PC1-2.png"
png(filename = file_name)
ggbiplot(fOTUsHMM_pca, var.axes = F, choices = 1:2)
dev.off()
ggbiplot(fOTUsHMM_pca, var.axes = F, choices = 1:2)
```


```{r}
prediction <- predict(fOTUsHMM_pca)[,1:20]
file_name <- "../visualisations/HMMsearch/heatmap_avg_20_first_PCs.png"
png(filename = file_name)
heatmap(as.matrix(prediction))
dev.off()
heatmap(as.matrix(prediction))

file_name <- "../visualisations/HMMsearch/heatmap_avg_400_first_fOTUs.png"
png(filename = file_name)
heatmap(as.matrix(fOTUsHMM[,-1])[,1:400])
dev.off()
heatmap(as.matrix(fOTUsHMM[,-1])[,1:400])
```

# NMDS

Let's perform non-metric multidimensional scaling:

```{r}
width <- dim(fOTUsHMM)[2]
ord_avg <- metaMDS(fOTUsHMM[,2:width])
```

# Save/Load R data

Save or load the data once again to speed-up things.

```{r eval=FALSE, include=TRUE}
#saveRDS(ord_avg, file = "../analyses/HMMsearch/R_data/avg_ord.rds")
ord_avg <- readRDS(file = "../analyses/HMMsearch/R_data/avg_ord.rds")
```

How well does the 2-D representation of multidimensional space represent the actual multidimensional space is done by using statistic called stress. Stress < 0.1 indicates that the 2-D representation is a good repreresentation of the multi-D space. Let's check that out now:

```{r}
ord_avg$stress
```

This is quite low which is great! Let's now plot out the relationship between the ordination dissimilarities and the distances in the ordination space:

```{r}
stressplot(ord_avg)
```

Observed dissimilarity of 1 means the fOTU doesn't have any HMMs incommon with the others.

Let's plot out the ordination space in 2-d:

```{r}
ordiplot(ord_avg)
points(ord_avg)
```

and let's filter away the outliers and make new plot from above:

```{r}
NMDS_points1 <- as_tibble(ord_avg$points) %>%
  filter(MDS1 > -10)

title <- "NMDS, MDS1 > -10"
qplot(MDS1, 
      MDS2, 
      data = NMDS_points1, 
      geom = c("point", "density2d")) + ggtitle(title)

path <- "../visualisations/HMMsearch/nmds_hmmsearch_avg_MDS1_greater_than_-10.png"

ggsave(filename = path,
       device = png, 
       width = 1000, 
       height = 750, 
       units = "in",
       limitsize = FALSE,
       dpi = 320)
```

Let's filter away slightly more:

```{r}
NMDS_points2 <- as_tibble(ord_avg$points) %>%
  filter(MDS1 > -2)

title <- "NMDS, -2 < MDS1"
qplot(MDS1, 
      MDS2, 
      data = NMDS_points2, 
      geom = c("point", "density2d")) + ggtitle(title)

path <- "../visualisations/HMMsearch/nmds_hmmsearch_avg_MDS1_greater_than_-2.png"

ggsave(filename = path,
       device = png, 
       width = 1000, 
       height = 750, 
       units = "in",
       limitsize = FALSE,
       dpi = 320)
```



# Create lists of fOTUs of interest

## Preview PCA plots

These PCA plots can be used to visually determine potential groupings.

```{r}

PC12s <- tibble(fOTUs = names(fOTUsHMM_pca$x[,1]), 
              PC1 = unname(fOTUsHMM_pca$x[,1]), 
              PC2 = unname(fOTUsHMM_pca$x[,2]))

title <- "PCA of e-values of matches in fOTUs"
qplot(PC1, 
      PC2, 
      data = PC12s, 
      geom = c("point", "density2d")) + ggtitle(title)

ggsave(filename = "../visualisations/HMMsearch/pca_hmmsearch_avg_PC1-2.png",
       device = png, 
       width = 1000, 
       height = 750, 
       units = "in",
       limitsize = FALSE,
       dpi = 320)

PC23s <- tibble(fOTUs = names(fOTUsHMM_pca$x[,1]), 
              PC2 = unname(fOTUsHMM_pca$x[,2]), 
              PC3 = unname(fOTUsHMM_pca$x[,3]))

title <- "PCA of e-values of matches in fOTUs"
qplot(PC2, 
      PC3, 
      data = PC23s, 
      geom = c("point", "density2d")) + ggtitle(title)

ggsave(filename = "../visualisations/HMMsearch/pca_hmmsearch_avg_PC2-3.png",
       device = png, 
       width = 1000, 
       height = 750, 
       units = "in",
       limitsize = FALSE,
       dpi = 320)

PC34s <- tibble(fOTUs = names(fOTUsHMM_pca$x[,1]), 
              PC3 = unname(fOTUsHMM_pca$x[,3]), 
              PC4 = unname(fOTUsHMM_pca$x[,4]))

title <- "PCA of e-values of matches in fOTUs"
qplot(PC3, 
      PC4, 
      data = PC34s, 
      geom = c("point", "density2d")) + ggtitle(title)

ggsave(filename = "../visualisations/HMMsearch/pca_hmmsearch_avg_PC3-4.png",
       device = png, 
       width = 1000, 
       height = 750, 
       units = "in",
       limitsize = FALSE,
       dpi = 320)
```

## Create a file for each list

By comparing visually the PCA data, we can see that they are located essentially in same locations. Thus it is unnecessary to create new lists based on the average e-value points when there already are points based on the maximum e-values.


# Session info

```{r}
sessionInfo()
```


# References
