---
title: "Searching further on viral matching HMM profiles"
author: "LJM"
date: "`r Sys.Date()`"
link-citations: true
bibliography: bibliography.bib
biblio-style: "alpha" #https://www.overleaf.com/learn/latex/Bibtex_bibliography_styles
#csl: "../uppsala-universitet-institutionen-for-biologisk-grundutbildning.csl"
subtitle: "Studying further what matched the fOTUs for purposes of trying to validate the results"
subject: "Study further what matched the fOTUs for purposes of trying to validate the results."
keywords: 
  - bioinformatics
  - HMMer
  - eggNOG 
output:
  bookdown::pdf_document2:
    keep_tex: no
    latex_engine: xelatex
    #highlight: pygments #Other options: default, tango, kate, monochrome, espresso, zenburn, haddock, and textmate
    #theme: spacelab
    # https://pandoc.org/MANUAL.html
    pandoc_args: ["--top-level-division=section",
                  "-V", "documentclass=article",
                  "-V", "linkcolor=MidnightBlue",
                  "-V", "citecolor=Aquamarine",
                  "-V", "urlcolor=NavyBlue",
                  "-V", "toccolor=NavyBlue",
                  "-V", "pagestyle=headings"]
  bookdown::html_document2:
    highlight: tango
    theme: yeti
    split_by: none # only generate a single output page
    self_contained: TRUE
    toc: yes
    toc_float: 
      collapsed: TRUE
      smooth_scroll: TRUE
      print: FALSE
    code_folding: show
    number_sections: TRUE
    code_download: TRUE
    pandoc_args: ["--top-level-division=section",
              "-V", "documentclass=report"]
              
---

\clearpage

# Introduction

It would be nice to know further what viruses contributed to HMMs that matched to the fOTUs. Fortunately EggNOG database v 5.0 [@Huerta-Cepas2019] provides that information. Let's download the relevant file from EggNOG database and taxonomic data from NCBI and create one big tsv file with all that information. Thereafter, let's check which HMMs matched significantly (with $-log_{10}\text{(e-value)}$ of 2 or higher) to which fOTUs. 

## Load libraries

```{r message=FALSE, warning=FALSE}
library(tidyverse)
```

\clearpage

# Download from eggNOG database

## Download members information on all viral HMMs

```{bash eval=FALSE, include=TRUE}
OUTPUT="../analyses/HMMsearch/validation_and_further_analyses/10239_members.tsv.gz"
wget --verbose --continue --output-document="$OUTPUT" \
"http://eggnog5.embl.de/download/eggnog_5.0/per_tax_level/10239/10239_members.tsv.gz"
gunzip "$OUTPUT"
```

## Download annotation information on all viral HMMs

```{bash eval=FALSE, include=TRUE}
OUTPUT="../analyses/HMMsearch/validation_and_further_analyses/10239_annotations.tsv.gz"
wget --verbose --continue --output-document="$OUTPUT" \
"http://eggnog5.embl.de/download/eggnog_5.0/per_tax_level/10239/10239_annotations.tsv.gz"
gunzip "$OUTPUT"
```

\clearpage

# Merge annotations and members files

```{bash eval=FALSE, include=TRUE}
INPUT1="../analyses/HMMsearch/validation_and_further_analyses/10239_annotations.tsv"
INPUT2="../analyses/HMMsearch/validation_and_further_analyses/10239_members.tsv"
SCRIPT="../scripts/merge_files.awk"
OUT="../analyses/HMMsearch/validation_and_further_analyses/10239_members_annotations.tsv"

awk -f $SCRIPT $INPUT1 $INPUT2 > $OUT
```

Now let's take a short look at the output

```{bash}
IN="../analyses/HMMsearch/validation_and_further_analyses/10239_members_annotations.tsv"
head -n 3 $IN
```

Looks pretty ok.

Now we need to fetch taxonomic information from NCBI based on mem_taxid column information and append it to this previous data.

\clearpage

# Fetch information from NCBI on each HMM

Let's use Biopython v. 1.75 [@Cock2009] to access NCBI's taxonomy database for retrieving interesting taxonomic information about the viruses which contributed their proteins in the HMMs used. Biopython comes also with a handy parser of the XML format data that is returned from NCBI on each request.

## Fetch taxonomic data

```{bash eval=FALSE, include=TRUE}
IN="../analyses/HMMsearch/validation_and_further_analyses/10239_members_annotations.tsv"
SCRIPT="../scripts/fetch_taxon_data.py"
AWK_PARSER="../scripts/tabinator.awk"
OUTPUT="../analyses/HMMsearch/validation_and_further_analyses/lineages.tsv"
# Retain fields of interest with AWK, process with python by augmenting the data with
# taxon info from NCBI and print out in a nicely machine readable format with AWK
tail -n+2 $IN | awk -F"\t" 'BEGIN{OFS=","}{print $2,$5}' |\
python3 $SCRIPT | awk -F"@" -f $AWK_PARSER > $OUTPUT
```

In this returned file `/analyses/HMMsearch/validation_and_further_analyses/lineages.tsv` there are some issues that should be fixed before it will be read with R and processed further. The issues are:

1. There are unnecessary spaces after `;` in lineages column, e.g. `Viruses; Caudovirales; Myoviridae; unclassified Myoviridae,Viruses; Caudovirales; Myoviridae; unclassified Myoviridae`.

2. For 350 records root taxon was downloaded as the first taxon. This is of no interest for us so we need to get rid of it. This task involves not only removing `root` from scientific names (2nd field) but all NAs in other taxonomic columns and leading commas in lineages fields (the 13th one).

3. For 350 other records generic Bacteria taxon was downloaded as the first taxon. This seems to be some type of incorrect relic and not a valid contributor organism for the HMMs. This task involves not only removing `Bacteria` from scientific names (2nd field) but all NAs in other taxonomic columns except `no_rank` (9th field) where `cellular organisms` should be removed and finally in lineages field (the 13th one) removing the leading `cellular organisms,` text.

Let's fix these issues in this order before we move on.

## Remove spaces from lineages field

```{bash eval=FALSE, include=TRUE}
IN="../analyses/HMMsearch/validation_and_further_analyses/lineages.tsv"
OUT="../analyses/HMMsearch/validation_and_further_analyses/taxon_data.tsv"
AWK_SCRIPT="../scripts/linDeWhitespacer.awk"
awk -F"\t" -f $AWK_SCRIPT $IN > $OUT
#sed -n '3'p $OUT
#tail -n 100 $TEMP | awk -F"\t" -f $AWK_SCRIPT | awk -F"\t" '{print $13}'
```

## Remove `root` taxons from some records

```{bash eval=FALSE, include=TRUE}
export AWKPATH="../scripts/awk_lib"
FILE="../analyses/HMMsearch/validation_and_further_analyses/taxon_data.tsv"
TEMP="../analyses/HMMsearch/validation_and_further_analyses/temp.tsv"
AWK_SCRIPT="../scripts/removeRoots.awk"
cp $FILE $TEMP
awk -F"\t" -f $AWK_SCRIPT $TEMP > $FILE 
rm $TEMP
```

## Remove `Bacteria` taxons from some records

```{bash eval=FALSE, include=TRUE}
FILE="../analyses/HMMsearch/validation_and_further_analyses/taxon_data.tsv"
TEMP="../analyses/HMMsearch/validation_and_further_analyses/temp.tsv"
AWK_SCRIPT="../scripts/removeBacteria.awk"
cp $FILE $TEMP
awk -F"\t" -f $AWK_SCRIPT $TEMP > $FILE
#sed -n '19,22'p $TEMP | awk -F"\t" -f $AWK_SCRIPT
rm $TEMP
```

## The output file

The retrieved data looks now the following: (the lines have been stripped so it looks nicer in `pdf`/`html`-formats)

```{bash}
FILE="../analyses/HMMsearch/validation_and_further_analyses/taxon_data.tsv"
head -n 5 $FILE | cut -c-83
```

\clearpage

# Augment existing data with taxonomic information

## Read in the data

```{r eval=FALSE, include=TRUE}
# The files where the data is
taxon_data_file <- "../analyses/HMMsearch/validation_and_further_analyses/taxon_data.tsv"
hmm_data_file <- "../analyses/HMMsearch/validation_and_further_analyses/10239_members_annotations.tsv"
fOTUs_data_file <- "../analyses/HMMsearch/max_fOTUsHMM.csv"

# read in the data
hmm_data <- read_tsv(file = hmm_data_file,
                     col_names = TRUE,
                     col_types = c("fcfccc"), 
                     na = "NA")

taxon_data <- read_tsv(file = taxon_data_file,
                     col_names = c("HMM",
                                   "sci_names",
                                   "superkingdoms",
                                   "orders",
                                   "families",
                                   "subfamilies",
                                   "genuses",
                                   "species",
                                   "no_ranks",
                                   "phylums",
                                   "classes",
                                   "unclassified",
                                   "lineages"),
                     col_types = c("ccccccccccccc"), 
                     na = "NA")

columns <- paste("c", strrep("d",8315), sep = "")
fOTUsHMM <- read_csv(file = fOTUs_data_file,
                     col_names = TRUE, 
                     col_types = columns)

```

## Join the data into one tibble

```{r eval=FALSE, include=TRUE}
joined_data <- hmm_data %>%
  left_join(.,
            taxon_data,
            by = "HMM")
```

## Create summary

What we'd like to do now is create a tsv file for each fOTU such that it contains all the significant matches with HMMs as rows and the relevant HMM data gathered above joined to the rows.

First let's define some functions

```{r eval=FALSE, include=TRUE}
# This function returns FALSE for columns/vectors
# where all elements are NA
not_all_na <- function(x){
  !all(is.na(x))
}

# This function removes columns from the tibble where all the elements
# are NA and returns the reduced tibble
remove_na_cols <- function(this_tbl){
  filtered <- this_tbl %>%
    select_if(not_all_na)
}
```

Now we'll simply iterate over each row in the tibble with fOTUs and $-log_{10}(\text{e-val})$:s, then remove all the columns (HMMs) (in each row) which didn't have any matches and pivot the HMMs with some $-log_{10}(\text{e-val})$:s as rows, join all the previously gathered data to each HMM and finally write this data to a `tsv`-file with the current fOTU name as file name.

```{r eval=FALSE, include=TRUE}
path <- "../analyses/HMMsearch/validation_and_further_analyses/fOTUwise_taxon_HMM_data/"
for (row in 1:nrow(fOTUsHMM)) {
  # Scrape the current fOTU name for use when writing files
  fOTU_name <- as.character(fOTUsHMM[row,][1])
  # Create the file name with path
  file_name <- paste(path,fOTU_name,".tsv",sep = "")
  fOTUsHMM[row,] %>%
    remove_na_cols() %>%
    select(-fOTU_name) %>%
    gather("HMM","e-value") %>%
    left_join(.,
               joined_data,
               by = "HMM") %>%
    write_tsv(.,
              path = file_name,
              append = FALSE,
              col_names = TRUE)
}
```


\clearpage

# Session info

```{r}
sessionInfo()
```

\clearpage

# References